taxize - taxonomic search and retrieval in R

### 2013-06-21 inst/doc/taxize_withcode_plos.Rnw
adding section on aggregating
manuscript updated fixes issue #156

\newcolumntype{L}{>{\raggedright\arraybackslash}m{5cm}} % creates now column type to wrap text
\newcolumntype{B}{>{\raggedright\arraybackslash}m{3cm}} % creates now column type to wrap text
\newcolumntype{C}{>{\raggedright\arraybackslash}m{3.5cm}} % creates now column type to wrap text
\newcolumntype{D}{>{\raggedright\arraybackslash}m{2.5cm}} % creates now column type to wrap text
%% END MACROS SECTION

@@

<<loadtaxize, eval=FALSE>>=
library(taxize)
@

@@

<<resolvenames, eval=FALSE, cache=TRUE>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)
temp[,-c(1,4)]
@

@@

<<tnrs, eval=FALSE,cache=TRUE>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[,-c(5:7)]
@

@@

<<itisclass, eval=FALSE,cache=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

<<taxname, eval=FALSE,cache=TRUE>>=
tax_name(query="Helianthus annuus", get="family", db="itis")
tax_name(query="Helianthus annuus", get="family", db="ncbi")
@

@@

<<interactive, eval=FALSE, echo=TRUE, cache=TRUE>>=
get_tsn(searchterm="Elisabethae", searchtype="sciname")
@

@@

<<interactive_many, eval=FALSE, echo=TRUE, cache=TRUE>>=
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm=splist, searchtype="sciname")
@

@@

<<interactive_na, eval=FALSE, echo=TRUE, cache=TRUE>>=
get_uid(sciname=c("Chironomus riparius", "aaa vva"))
@

@@

<<downstream, eval=FALSE, cache=TRUE>>=
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=FALSE, cache=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank_one, eval=FALSE, cache=TRUE>>=
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)
@

@@

<<genbank_two, eval=FALSE, cache=TRUE>>=
out[grep("RAG1", out$genesavail, ignore.case=T),]
@

@@

>>>>

\newcolumntype{L}{>{\raggedright\arraybackslash}m{5cm}} % creates now column type to wrap text
%% END MACROS SECTION

@@

<<loadtaxize, eval=TRUE>>=
library(taxize)
@

@@

<<resolvenames, eval=TRUE, cache=TRUE>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)
temp[,-c(1,4)]
@

@@

<<tnrs, eval=TRUE,cache=TRUE>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[,-c(5:7)]
@

@@

<<itisclass, eval=TRUE,cache=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

<<taxname, eval=TRUE,cache=TRUE>>=
tax_name(query="Helianthus annuus", get="family", db="itis")
tax_name(query="Helianthus annuus", get="family", db="ncbi")
@

@@

<<interactive, eval=TRUE, echo=TRUE, cache=TRUE>>=
get_tsn(searchterm="Heliastes", searchtype="sciname")
@

@@

<<interactive_many, eval=TRUE, echo=TRUE, cache=TRUE>>=
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm=splist, searchtype="sciname")
@

@@

<<interactive_na, eval=TRUE, echo=TRUE, cache=TRUE>>=
get_uid(sciname=c("Chironomus riparius", "aaa vva"))
@

@@

<<downstream, eval=TRUE, cache=TRUE>>=
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=TRUE, cache=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank_one, eval=TRUE, cache=TRUE>>=
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)
@

@@

<<genbank_two, eval=TRUE, cache=TRUE>>=
out[grep("RAG1", out$genesavail, ignore.case=T),]
@

@@

\subsection*{Aggregating data by any taxonomic rank}

In biology, one can asks questions at varying taxonomic levels. Do species differ? Do genera differ? This use case is easily handled in taxize. A very basic function called \emph{rankagg} will aggregate any data to a specific taxonomic level. In this example, a single column \emph{abundance} is aggregated by any arithmetic function (e.g., mean) at any taxonomic rank (e.g., Family).

<<aggregate_abd, eval=TRUE, cache=TRUE>>=
library(vegan)
data(dune.taxon)
dat <- dune.taxon
dat$abundance <- round(rlnorm(n=nrow(dat),meanlog=5,sdlog=2),0)
head( rankagg(data=dat, datacol="abundance", rank="Genus") )
head( rankagg(data=dat, "abundance", rank="Family", fxn="mean") )
@

### 2013-06-21 inst/doc/taxize_withcode_plos.tex
manuscript updated fixes issue #156

\newcolumntype{L}{>{\raggedright\arraybackslash}m{5cm}} % creates now column type to wrap text
\newcolumntype{B}{>{\raggedright\arraybackslash}m{3cm}} % creates now column type to wrap text
\newcolumntype{C}{>{\raggedright\arraybackslash}m{3.5cm}} % creates now column type to wrap text
\newcolumntype{D}{>{\raggedright\arraybackslash}m{2.5cm}} % creates now column type to wrap text
%% END MACROS SECTION
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus"}, \hlstring{"Pinus contort"}, \hlstring{"Poa anua"}, \hlstring{"Abis magnifica"}, 
    \hlstring{"Rosa california"}, \hlstring{"Festuca arundinace"}, \hlstring{"Sorbus occidentalos"}, \hlstring{"Madia sateva"})
\hlfunctioncall{tnrs}(query = mynames)[, -\hlfunctioncall{c}(5:7)]
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(\hlfunctioncall{get_tsn}(specieslist, \hlstring{"sciname"}))
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"itis"})
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"ncbi"})
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Elisabethae"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
splist <- \hlfunctioncall{c}(\hlstring{"annona cherimola"}, \hlstring{"annona muricata"}, \hlstring{"quercus robur"}, \hlstring{"shorea robusta"}, 
    \hlstring{"pandanus patina"}, \hlstring{"oryza sativa"}, \hlstring{"durio zibethinus"})
\hlfunctioncall{get_tsn}(searchterm = splist, searchtype = \hlstring{"sciname"})
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
out[\hlfunctioncall{grep}(\hlstring{"RAG1"}, out$genesavail, ignore.case = T), ]
\end{alltt}
\end{kframe}
\end{knitrout}

@@

>>>>

\newcolumntype{L}{>{\raggedright\arraybackslash}m{5cm}} % creates now column type to wrap text
%% END MACROS SECTION
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
mynames <- \hlfunctioncall{c}(\hlstring{"Helianthus annuus"}, \hlstring{"Pinus contort"}, \hlstring{"Poa anua"}, \hlstring{"Abis magnifica"}, 
    \hlstring{"Rosa california"}, \hlstring{"Festuca arundinace"}, \hlstring{"Sorbus occidentalos"}, \hlstring{"Madia sateva"})
\hlfunctioncall{tnrs}(query = mynames)[, -\hlfunctioncall{c}(5:7)]
\end{alltt}
\begin{verbatim}
        submittedName        acceptedName    sourceId score
7   Helianthus annuus   Helianthus annuus iPlant_TNRS  1.00
4       Pinus contort      Pinus contorta iPlant_TNRS  0.98
5            Poa anua           Poa annua iPlant_TNRS  0.96
3      Abis magnifica     Abies magnifica iPlant_TNRS  0.96
8     Rosa california    Rosa californica iPlant_TNRS  0.99
2  Festuca arundinace Festuca arundinacea iPlant_TNRS  0.99
1 Sorbus occidentalos Sorbus occidentalis iPlant_TNRS  0.99
6        Madia sateva        Madia sativa iPlant_TNRS  0.97
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
specieslist <- \hlfunctioncall{c}(\hlstring{"Abies procera"}, \hlstring{"Pinus contorta"})
\hlfunctioncall{classification}(\hlfunctioncall{get_tsn}(specieslist, \hlstring{"sciname"}))
\end{alltt}
\begin{verbatim}
[[1]]
        parentName parentTsn      rankName       taxonName    tsn
1                                  Kingdom         Plantae 202422
2          Plantae    202422    Subkingdom  Viridaeplantae 846492
3   Viridaeplantae    846492  Infrakingdom    Streptophyta 846494
4     Streptophyta    846494      Division    Tracheophyta 846496
5     Tracheophyta    846496   Subdivision Spermatophytina 846504
6  Spermatophytina    846504 Infradivision    Gymnospermae 846506
7     Gymnospermae    846506         Class       Pinopsida 500009
8        Pinopsida    500009         Order         Pinales 500028
9          Pinales    500028        Family        Pinaceae  18030
10        Pinaceae     18030         Genus           Abies  18031
11           Abies     18031       Species   Abies procera 181835

[[2]]
        parentName parentTsn      rankName       taxonName    tsn
1                                  Kingdom         Plantae 202422
2          Plantae    202422    Subkingdom  Viridaeplantae 846492
3   Viridaeplantae    846492  Infrakingdom    Streptophyta 846494
4     Streptophyta    846494      Division    Tracheophyta 846496
5     Tracheophyta    846496   Subdivision Spermatophytina 846504
6  Spermatophytina    846504 Infradivision    Gymnospermae 846506
7     Gymnospermae    846506         Class       Pinopsida 500009
8        Pinopsida    500009         Order         Pinales 500028
9          Pinales    500028        Family        Pinaceae  18030
10        Pinaceae     18030         Genus           Pinus  18035
11           Pinus     18035       Species  Pinus contorta 183327
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"itis"})
\end{alltt}
\begin{verbatim}
          V1
1 Asteraceae
\end{verbatim}
\begin{alltt}
\hlfunctioncall{tax_name}(query = \hlstring{"Helianthus annuus"}, get = \hlstring{"family"}, db = \hlstring{"ncbi"})
\end{alltt}
\begin{verbatim}
      family
1 Asteraceae
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_tsn}(searchterm = \hlstring{"Heliastes"}, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
           combinedname    tsn
1     Heliastes bicolor 615238
2   Heliastes chrysurus 615250
3     Heliastes cinctus 615573
4  Heliastes dimidiatus 615257
5  Heliastes hypsilepis 615273
6 Heliastes immaculatus 615639
7 Heliastes opercularis 615300
8      Heliastes ovalis 615301
 1 
NA 
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
splist <- \hlfunctioncall{c}(\hlstring{"annona cherimola"}, \hlstring{"annona muricata"}, \hlstring{"quercus robur"}, \hlstring{"shorea robusta"}, 
    \hlstring{"pandanus patina"}, \hlstring{"oryza sativa"}, \hlstring{"durio zibethinus"})
\hlfunctioncall{get_tsn}(searchterm = splist, searchtype = \hlstring{"sciname"})
\end{alltt}
\begin{verbatim}
[1] "506198" "18098"  "19405"  "506787" "507376" "41976"  "506099"
attr(,"class")
[1] "tsn"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{get_uid}(sciname = \hlfunctioncall{c}(\hlstring{"Chironomus riparius"}, \hlstring{"aaa vva"}))
\end{alltt}
\begin{verbatim}
[1] "315576" NA      
attr(,"class")
[1] "uid"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{col_downstream}(name = \hlstring{"Apis"}, downto = \hlstring{"Species"})[[1]]
\end{alltt}
\begin{verbatim}
  childtaxa_id     childtaxa_name childtaxa_rank
1      6971712 Apis andreniformis        Species
2      6971713        Apis cerana        Species
3      6971714       Apis dorsata        Species
4      6971715        Apis florea        Species
5      6971716 Apis koschevnikovi        Species
6      6845885     Apis mellifera        Species
7      6971717   Apis nigrocincta        Species
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
out[\hlfunctioncall{grep}(\hlstring{"RAG1"}, out$genesavail, ignore.case = T), ]
\end{alltt}
\begin{verbatim}
        spused length
413 Umbra limi    732
427 Umbra limi    959
434 Umbra limi   1631
                                                                            genesavail
413 isolate UlimA recombinase activating protein 1 (rag1) gene, exon 3 and partial cds
427           recombination-activating protein 1 (RAG1) gene, intron 2 and partial cds
434                        recombination-activating protein 1 (RAG1) gene, partial cds
    access_num       ids
413   JX190826 394772608
427   AY459526  45479841
434   AY380548  38858304
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\subsection*{Aggregating data by any taxonomic rank}

In biology, one can asks questions at varying taxonomic levels. Do species differ? Do genera differ? This use case is easily handled in taxize. A very basic function called \emph{rankagg} will aggregate any data to a specific taxonomic level. In this example, a single column \emph{abundance} is aggregated by any arithmetic function (e.g., mean) at any taxonomic rank (e.g., Family).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlfunctioncall{library}(vegan)
\hlfunctioncall{data}(dune.taxon)
dat <- dune.taxon
dat$abundance <- \hlfunctioncall{round}(\hlfunctioncall{rlnorm}(n = \hlfunctioncall{nrow}(dat), meanlog = 5, sdlog = 2), 0)
\hlfunctioncall{head}(\hlfunctioncall{rankagg}(data = dat, datacol = \hlstring{"abundance"}, rank = \hlstring{"Genus"}))
\end{alltt}
\begin{verbatim}
       Genus Result
1     Bellis    347
2   Empetrum   1169
3     Juncus    411
4       Aira      5
5 Eleocharis    158
6      Rumex      2
\end{verbatim}
\begin{alltt}
\hlfunctioncall{head}(\hlfunctioncall{rankagg}(data = dat, \hlstring{"abundance"}, rank = \hlstring{"Family"}, fxn = \hlstring{"mean"}))
\end{alltt}
\begin{verbatim}
        Family Result
1   Asteraceae  289.6
2  Empetraceae 1169.0
3    Juncaceae  205.5
4      Poaceae  452.2
5   Cyperaceae  158.0
6 Polygonaceae    2.0
\end{verbatim}
\end{kframe}
\end{knitrout}
